Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-06-03 paper/paper.tex
revision of paper based on initial reviews (#36)
* update all existing figures (new figs and text revisions not yet added)

* add new figures (not linked to text / legends yet)

* revise first two results sections

* add dn/ds and comparator predictions

* revision of paper

* generate track-changes manuscript

* finish revising paper

* updated track changes manuscript

\subsection*{Mutation-effect estimates are robust to subsampling and phylogeny, with some evidence of epistasis in spike}

@@

The modest variation in estimates from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real shifts in mutation effects during SARS-CoV-2 evolution~\citep{starr2022shifting,moulana2022compensatory}.
To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
Correlations increased with this count threshold, consistent with reduced statistical noise for larger mutation counts.
But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).

@@

An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations even after aggregating across clades.
To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
%We also note that the fact that two trees (which have independently inferred topologies) give nearly identical fitness estimates indicates that uncertainty in tree topology has minimal effect.
% The insensitivity of our fitness estimates to uncertainty in tree topology is expected: the main uncertainty in densely sampled viral phylogenies involves polytomies of closely related sequences, but our estimates only use the counts of mutations along tree branches, which are unaffected by uncertainty in the resolution of polytomies of closely related sequences.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figs/dist.pdf}
\caption{
Distribution of effects of different classes of mutations.
{\bf (A)}
Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
Neutral mutations have effects of zero (dashed gray vertical lines), and deleterious mutations have negative effects.
{\bf (B)}
Effects of each class of mutation for each viral gene.
Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
Mutation types are color-coded as in panel A.
The apparent constraint on synonymous mutation in ORF9b is probably because this gene is encoded in an overlapping reading frame with N~\cite{jungreis2021conflicting}.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
\label{fig:dist}
}
\end{figure*}

@@

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
Observations such as these could help guide experimental studies to better understand protein function.

@@

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio. 
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

@@

>>>>

\subsection*{Mutation-effect estimates are robust to subsampling, with some evidence of epistasis in spike}

@@

The modest variation in estimates from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real shifts in mutation effects during SARS-CoV-2 evolution~\citep{starr2022shifting,moulana2022compensatory}.
To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
Correlations increased with this count threshold, consistent with reduced statistical noise for larger mutation counts.
But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
In particular, the interactive version of Figure~\ref{fig:corr}A shows that mutations that are more beneficial in Omicron BA.5 than Delta are often antibody-escape mutations (eg, K444N or G446S in spike~\citep{greaney2022antibody})---a result that makes sense, since newer variants like Omicron BA.5 are evolving under increased immune selection compared to earlier variants like Delta that circulated in a more immunologically naive population.

@@

An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations even after aggregating across clades.
To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figs/dist.pdf}
\caption{
Distribution of effects of different classes of mutations.
{\bf (A)}
Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
Neutral mutations have effects of zero (dashed gray vertical lines), and deleterious mutations have negative effects.
{\bf (B)}
Effects of each class of mutation for each viral gene.
Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
Mutation types are color-coded as in panel A.
The apparent constraint on synonymous mutation in ORF9b is probably because this gene is encoded in an overlapping reading frame with N~\cite{jungreis2021conflicting}.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
See Figure~\ref{fig:effects_dist_position_order} for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
\label{fig:dist}
}
\end{figure*}

@@

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}. 
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
Observations such as these could help guide experimental studies to better understand protein function.

@@

\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not,yang2000estimating}.
A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio. 
For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.

@@

\begin{figure*}
\includegraphics[width=\linewidth]{figs/effects_dist_position_order.pdf}
\caption{
A version of Figure~\ref{fig:dist}B but with genes ordered by position in the genome rather than extent of constraint on nonsynonymous mutations.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist_position_order.html} for an interactive version of this plot.
\label{fig:effects_dist_position_order}
}
\end{figure*}
